---
title: "Lab 10: t-tests"
format: 
  html:
    toc-depth: 3
---

## Learning Objectives

## Getting started

Before you begin the lab activities, review the steps for setting up a project for a lab activity:

-   [Weekly Lab Setup](../../resources/weekly-lab-setup.qmd)

Then, complete the following steps to make sure you are fully set up.

1.  **Get the Lab Worksheet.**

    Pick up a physical copy of the lab worksheet, or print one if you are working outside of class.\
    <a href="lab-10-worksheet.pdf" download class="btn btn-outline-primary mt-3">Download Lab Worksheet (PDF, if needed)</a>

2.  **Open Posit Cloud and create Lab 10 in Your Workspace.**

3.  Create a new R script and save it as **lab-10-script.R**.

4.  Add the following code to your R script and save the script. Upon save, you should see a message in the source prompting you to install the package. Click "Install". Then run the line of code.

    ```{r}
    #| label: load-packages
    #| echo: true

    # load packages --------------------------------------------------------------

    library(tidyverse)
    library(haven)
    ```

::: callout-note
## Checkpoint

At this point, you should have:

-   These instructions open in a web browser.
-   Your Lab 10 project open in Posit Cloud in another browser window.
-   The required packages installed and loaded without errors
-   The Lab 10 worksheet in front of you.

Do not continue until all of the above steps are working correctly.
:::

## Overview

In this lab, you will analyze the relationship between two categorical variables using **contingency analysis**. A contingency table summarizes the joint distribution of two categorical variables and allows us to estimate measures of association such as **relative risk (RR)** and **odds ratio (OR)**. These measures describe the magnitude and direction of an association between two binary variables. We will also conduct a **chi-square test of independence** to determine whether the observed association is statistically significant.

You will use data from the National Health and Nutrition Examination Survey (NHANES), a large, nationally representative health survey conducted in the United States. For this lab, you will examine the relationship between sleep duration and depressive symptoms among adults.

## Prerequisites

In your R script, load the following packages:

```{r}

library(tidyverse) # for data manipulation and visualization
library(haven) # for reading SAS XPT files
library(palmerpenguins) # for penguin data
library(lterdatasampler) # for bison data
library(PairedData) # for shoulder data

```

## Sleep duration: one-sample test

In this activity, you will:

-    read in NHANES sleep data and keep the first 100 observations
-    visualize the data using a histogram
-    assess the distribution to decide between a t-test and a Wilcoxon test
-    test whether mean (or median) sleep duration differs from 8 hours
-    interpret the result in plain language

Code

```{r}
#| echo:true
slq_data <-
  read_xpt("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/SLQ_L.xpt") |> 
  drop_na(SLD012) |>
  slice_head(n = 100)

slq_data |>
  ggplot(aes(x = SLD012)) +
  geom_histogram(binwidth = 1, boundary = 0) +
  labs(
    x = "Hours of sleep",
    y = "Frequency",
    title = "Distribution of sleep duration"
  ) +
  theme_minimal()
```

## Bison body mass: two-sample comparison

In this activity, you will:

-    read in the bison dataset and keep 100 observations for each sex
-    visualize body mass distributions for males and females using histograms
-    assess the distributions to decide between a t-test and a Mann–Whitney test
-    compare body mass between sexes using either `t.test()` or `wilcox.test()`
-    interpret the results in terms of differences between groups

Use the following code:

```{r}
#| echo: true
bison_data <-
  knz_bison |> 
  drop_na(animal_sex, animal_weight) |>
  slice_head(n = 100, by = animal_sex)

bison_data |>
  ggplot(aes(x = animal_weight)) +
  geom_histogram(binwidth = 50) +
  facet_wrap(~animal_sex, ncol = 1)
```

In this activity, you will work with a dataset of shoulder strength measurements.

-    each row represents an individual
-    variables include **Left** and **Right** shoulder strength measurements
-    individuals are classified as **Swimmer** or **Control (non-swimmer)**
-    the goal is to compare strength **within individuals**, not between individuals

You will:

-    calculate **strong arm** and **weak arm** for each individual
-    note that comparing **left vs right directly is misleading** because people are left- or right-handed
-    instead, compute **strong arm − weak arm** to create a meaningful paired comparison
-    perform **paired t-tests** comparing strong vs weak arm strength
-    run separate tests for:
    -    swimmers
    -    non-swimmers
-    optionally, test **left vs right in non-swimmers** to confirm the difference is near zero
-    interpret results in terms of within-individual differences

### Code

```{r}
#| echo: true

data(Shoulder)

shoulder_data <- 
  tibble(Shoulder) |> 
  rowwise() |>
  mutate(
    strongarm = max(Left, Right),
    weakarm = min(Left, Right)
  ) |>
  ungroup()

shoulder_data |>
  ggplot(aes(x = strongarm)) +
  geom_histogram(bins = 10)
  
shoulder_data |>
  ggplot(aes(x = weakarm)) +
  geom_histogram(bins = 10)

swimmers <- 
  shoulder_data |> 
  filter(Group == "Swimmer") |>
  print()

nonswimmers <-
  shoulder_data |> 
  filter(Group == "Control") |>
  print()

```

## Wrap-up and Submission

1.  Make sure your script is saved in your project on Posit Cloud.
2.  Ask the instructor to explain anything you aren't sure about.
3.  **Show your handout to a Learning Assistant for a completion grade before you leave lab**. You may do this as soon as you finish. Keep the handout for yourself.
