---
title: "How to Export a Graph from RStudio for a Poster"
description: "A practical guide to making poster-ready figures in R with ggplot2 and exporting them at fixed dimensions using ggsave()."
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

When creating a poster in PowerPoint, figures should be designed and exported from R at their final size rather than resized later.

This requires a deliberate workflow that controls both the appearance of the plot and how it is saved.

-   Create figures using `ggplot2`
-   Use a large base font size suitable for posters
-   Export figures with `ggsave()` using fixed dimensions (in inches)
-   Insert into PowerPoint without major resizing

This produces figures that are readable, properly scaled, and consistent across the poster.

## The Problem

A common issue is getting a figure from R into PowerPoint that is actually readable on a poster.

Many students use approaches that seem convenient but produce poor results:

-   copying a figure from the *Plots* tab in RStudio
-   using the *Export* button in the *Plots* tab adjusting image size and font size
-   taking a screenshot of the graph

These approaches do not preserve control over:

-   figure size
-   text size
-   resolution
-   aspect ratio (length to width ratio)

As a result, when the figure is placed into PowerPoint and resized:

-   text becomes too small to read
-   images appear blurry or pixelated
-   plots become stretched or distorted
-   facet panels become cramped
-   figures look inconsistent across the poster

## General Approach

The solution is to design and export figures in R at the exact size they will appear in PowerPoint.

This means:

-   deciding how much space the figure will occupy (e.g., half-column or full-column)
-   setting text size appropriate for poster viewing
-   exporting the image with fixed width and height in inches
-   avoiding major resizing after insertion into PowerPoint

This approach ensures:

-   readable text
-   clean image quality
-   correct proportions
-   consistent layout across figures

## Recommended workflow

### 1. Create and store the graph

-   Use `ggplot2` to build your figure
-   Assign the plot to an object using a descriptive name (e.g., `penguin_scatter`)
-   Build the full plot before saving (do not rely on the last displayed plot)
-   Use clear, consistent naming for figure objects
-   Preview the plot in RStudio to check appearance before exporting

### 2. Choose a theme and set `base_size = 20`

-   Start with your existing plot object (e.g., `penguin_scatter`)
-   Add a theme layer with an appropriate base size, and assign the result to a new object (e.g., `fig1`)
    -   Example: `fig1 <- penguin_scatter + theme_minimal(base_size = 20)`
-   This ensures:
    -   the plot is readable in the RStudio *Plots* tab
    -   the saved image will also have appropriately sized text for a poster
-   Recommended theme options:
    -   `theme_minimal(base_size = 20)`
    -   `theme_bw(base_size = 20)`
    -   `theme_classic(base_size = 20)`
    -   `theme_light(base_size = 20)`
-   The `base_size` setting controls most text, including:
    -   axis titles and labels
    -   legend text
    -   facet labels
-   Adjust other visual elements directly in geoms as needed:
    -   point size → `geom_point(size = ...)`
    -   line width → `geom_line(linewidth = ...)`
    -   text labels → `geom_text(size = ...)`, `geom_label(size = ...)`

### 3. Export with `ggsave()`

-   Use `ggsave()` to save your figure as a PNG file\
-   Include the following arguments:
    -   `filename =` (name of the output file, e.g., `"penguin_scatter.png"`)
    -   `plot =` (the plot object, e.g., `fig1`)
    -   `width =` (in inches)
    -   `height =` (in inches)
    -   `units = "in"`
-   Choose width based on how the figure will be used in PowerPoint:
    -   full-column figure: `width = 13.5`
    -   half-column figure: `width = 6` to `7`
-   Set height intentionally based on the content:
    -   adjust to maintain good proportions
    -   ensure text, points, and facets are not crowded
    -   there is no single correct value; expect to iterate
-   Use PNG format for clean, high-quality images suitable for posters

### 4. Open and inspect the saved file

-   Go to the *Files* tab in RStudio
-   Locate the exported image file (e.g., `"penguin_scatter.png"`)
-   Click the file to open it
-   Check the figure carefully:
    -   text size is readable
    -   points and lines are appropriately sized
    -   nothing looks crowded or cut off
    -   proportions look correct (not stretched)
-   If the figure does not look right:
    -   adjust width and/or height in `ggsave()`
    -   revise theme or geom sizes as needed
    -   re-save and check again

### 5. Insert the image into PowerPoint

-   In RStudio, go to the *Files* tab
-   Check the box next to your image file (e.g., `"penguin_scatter.png"`)
-   Click **More → Export...**
-   Save the file (typically to your *Downloads* folder)
-   In PowerPoint:
    -   Go to **Insert → Pictures → This Device**
    -   Navigate to your *Downloads* folder
    -   Select the image and insert it
-   After inserting:
    -   avoid large resizing (especially enlarging)
    -   position the image in its intended space on the poster

## Examples

The example below uses the `penguins` dataset from the `palmerpenguins` package.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)

penguin_scatter <- 
  penguins |>
  drop_na() |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Penguin body mass vs. flipper length",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_bw()

penguin_scatter
```

### Example 1: Full-column figure

This example saves a figure that takes up the entire width of a poster column, 13.5 inches.

Setting `base_size = 20` yields a good font size for this image.

```{r}
#| label: penguin-full-column
#| eval: false

fig1 <- 
  penguin_scatter +
  theme_bw(base_size = 20)

ggsave(
  filename = "fig1.png",
  plot = fig1,
  width = 13.5,
  height = 9,
  units = "in"
)
```

### Example 2: Half-column figure

This version uses the same graph but creates a smaller image suitable for half of a poster column.

Because the image is smaller, there isn't as much room for text on the figure. You can alleviate this problem by moving the legend to the bottom of the graph.

```{r}
#| label: penguin-half-column
#| fig-width: 4
#| fig-height: 4

penguin_scatter_half <-
  penguin_scatter +
  theme(legend.position = "bottom")

penguin_scatter_half
```

Then save the image using `ggsave()`.

-   Make the base font size smaller when saving the image (e.g. 16)
-   Note: The `theme()` layer is repeated here because adding `theme_bw()` resets any theme elements that had been customized in the previous step.

```{r}
#| label: penguin-half-export
#| eval: false

fig2 <- 
  penguin_scatter_half +
  theme_bw(base_size = 16) +
  theme(legend.position = "bottom")

ggsave(
  filename = "fig2.png", 
  plot = fig2, 
  width = 6.5, 
  height = 5, 
  units = "in"
)
```

## Facetted graphs

-   Facetted plots require additional layout control to remain readable on a poster
-   Use `facet_wrap()` with the `ncol` argument to control the number of columns
-   Adjust `ncol` based on the width of the figure:
    -   half-column figure:
        -   use `ncol = 1` or `ncol = 2`
    -   full-column figure:
        -   use `ncol = 3` or more
-   Choose a layout that:
    -   keeps each panel large enough to read
    -   avoids overcrowding
    -   makes efficient use of available space
-   The best choice depends on:
    -   the type of plot
    -   the number of facets
    -   how much detail each panel contains
-   Expect to adjust both `ncol` and figure height to get a clean result

### Example 3: Facetted figure

This example uses facets and adjusts the facet layout with `ncol`.

Note that the color legend can be removed because it is now redundant with the facet labels.

```{r}
#| label: facetted-setup
#| message: false
#| warning: false
#| fig-asp: 0.45
#| eval: true
#| 
penguin_scatter_facetted <- 
  penguin_scatter +
  facet_wrap(~species, ncol = 3) +
  guides(color = "none")

penguin_scatter_facetted
```

To export that image for a full column width:

```{r}
#| label: penguin-facet-full
#| eval: false

fig3 <- 
  penguin_scatter_facetted +
  theme_bw(base_size = 20)

ggsave(
  filename = "fig3.png",
  plot = fig3,
  width = 13.5,
  height = 6,
  units = "in"
)
```

### Example 4: Facetted figure with vertical layout

For a narrower version of the same facetted graph, you might change the facet layout to `ncol = 1`. This creates a vertical stack of panels that can be saved as a half-column figure.

Like the half-column example above, the resulting figure will have a title that is too large and a color legend that takes up too much space on the side of the graph. Rather than making the base font size smaller on this graph (its still a large, featured graph, just vertically oriented), we can tweak some other elements:

-   Keep the title from being too long, by giving the plot a new title using `labs()`:

    -   Make the title shorter, e.g. `title = "Penguin Measurements"`

    -   Move info to an optional subtitle, e.g. `subtitle = "Body mass vs. flipper length"`

    -   Keep the original title, but put in a manual line break `\n`, e.g. `title = "Penguin body mass vs.\nflipper length"`

-   Move the legend to the bottom of the graph, and style is so it looks nice there, using `theme()`:

    -   Move it to the bottom with `legend.position = "bottom"`

    -   Put the legend title above the legend with `legend.title.position = "top"`

    -   Center the legend title with `legend.title = element_text(hjust = .5)`

Here is the figure with a single column of facets, a wrapped title, and a repositioned legend:

```{r}
#| label: facetted-long-setup
#| message: false
#| warning: false
#| fig-width: 4
#| fig-asp: 2.25

penguin_scatter_facetted_long <- 
  penguin_scatter +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Penguin body mass vs.\nflipper length") +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    legend.title = element_text(hjust = .5)
  )

penguin_scatter_facetted_long
```

To export this version for a half-column width:

```{r}
#| label: penguin-half-column-facet 
#| eval: false 

fig4 <- 
  penguin_scatter_facetted_long +
  theme_bw(base_size = 20) +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    legend.title = element_text(hjust = .5)
  )

ggsave(
  filename = "fig4.png",
  plot = fig4,
  width = 6,
  height = 13.5,
  units = "in"
)
```
