---
title: |
  | Lecture 15
  | Comparing Two Means
subtitle: "ABD 3e Chapter 12"
description: ""
format:
  revealjs:
    fig-width: 8
    fig-height: 6
editor: 
  markdown: 
    wrap: 72
---
## Comparing Two Means

::::: columns
::: {.column width="50%"}
-   Many biological questions involve comparing the means of two groups.
-   Examples include treatment versus control or habitat comparisons.
-   Different study designs require different versions of the t test.
:::

::: {.column width="50%"}
```{r}
#| label: fig-penguin-bill-length-comparison
#| fig-cap: Distribution of bill length for Adelie and Gentoo penguins. Data from the Palmer Penguins dataset.
#| fig-alt: Two stacked histograms showing bill length distributions for Adelie and Gentoo penguins. Each panel shows counts of bill length measurements in millimeters.

library(dplyr)
library(ggplot2)
library(palmerpenguins)

penguins |> 
  tidyr::drop_na(bill_length_mm) |> 
  filter(species %in% c("Adelie", "Gentoo")) |> 
  ggplot(
    aes(x = bill_length_mm, fill = species)
  ) +
  geom_histogram(
    bins = 25,
    color = "white",
    linewidth = 0.3
  ) +
  geom_vline(
    aes(xintercept = bill_length_mm),
    ~ summarize(
      .x,
      bill_length_mm = mean(bill_length_mm), 
      .by = species
    ),
    color = "red",
    linewidth = 1.5
  ) +
  facet_wrap(~ species, ncol = 1) +
  see::scale_fill_oi(guide = "none") +
  theme_minimal(base_size = 18) +
  labs(
    x = "Bill length (mm)",
    y = "Count",
  )
```
:::
:::::

## The [independent two-sample t test]{.keyword}

::::: columns
::: {.column width="50%"}
1.  Used when observations in the two groups are independent.
2.  The test evaluates whether the difference between group means is
    greater than expected by chance.
3.  Common example is a treatment group compared with a control group.
:::

::: {.column width="50%"}
```{r}
#| label: fig-penguin-bill-length-comparison-means
#| fig-cap: Distribution of bill length for Adelie and Gentoo penguins. Red vertical lines mark the sample mean bill length for each species. Data from the Palmer Penguins dataset.
#| fig-alt: Two stacked histograms showing bill length distributions for Adelie and Gentoo penguins. Each panel shows counts of bill length measurements in millimeters, with a red vertical line indicating the mean bill length for that species.

penguins |> 
  tidyr::drop_na(bill_length_mm) |> 
  filter(species %in% c("Adelie", "Gentoo")) |> 
  ggplot(
    aes(x = bill_length_mm, fill = species)
  ) +
  geom_histogram(
    bins = 25,
    color = "white",
    linewidth = 0.3
  ) +
  geom_vline(
    aes(xintercept = bill_length_mm),
    ~ summarize(
      .x,
      bill_length_mm = mean(bill_length_mm), 
      .by = species
    ),
    color = "red",
    linewidth = 1.5
  ) +
  facet_wrap(~ species, ncol = 1) +
  see::scale_fill_oi(guide = "none") +
  theme_minimal(base_size = 18) +
  labs(
    x = "Bill length (mm)",
    y = "Count",
  )
```
:::
:::::

## The [paired t test]{.keyword}

::::: columns
::: {.column width="50%"}
1.  Used when observations occur in natural pairs such as before–after
    measurements.
2.  The analysis focuses on the differences within each pair.
3.  The mean of these differences is tested against zero.
:::

::: {.column width="50%"}
![Paired measurements connected by lines illustrating before and after
observations. Source: Whitlock & Schluter
3e.](images/paired-t-test.png){#fig-paired-t-test
fig-alt="Pairs of points connected by lines representing before and after measurements for the same subjects."}
:::
:::::

## Choosing the Correct Test

::::: columns
::: {.column width="50%"}
1.  The appropriate t test depends on the study design.
2.  One-sample tests compare a mean to a hypothesized value.
3.  Two-sample and paired tests compare means between groups.
:::

::: {.column width="50%"}
![Decision tree for selecting one-sample, two-sample, or paired t tests.
Source: Whitlock & Schluter
3e.](images/t-test-decision-tree.png){#fig-t-test-decision-tree
fig-alt="Flowchart guiding the choice between one-sample, independent two-sample, and paired t tests."}
:::
:::::