---
title: |
  | Lecture 18
  | Designing Experiments
subtitle: "ABD 3e Chapter 14"
description: "Introduces how well-designed experiments enable causal inference by reducing bias and sampling error, while contrasting their strengths with observational studies and outlining practical strategies for study design and sample size planning."
format:
  revealjs:
    fig-width: 8
    fig-height: 6
editor: 
  markdown: 
    wrap: 72
code-annotations: select
---

## Learning Objectives

-   Distinguish between association and causation and explain why
    observational studies cannot establish causation
-   Define confounding and explain how it biases inference
-   Describe how controls, random assignment, and blinding reduce bias
    in experiments
-   Identify experimental units and distinguish true replication from
    pseudoreplication
-   Explain how replication, balance, and blocking reduce sampling error
    and improve precision
-   Interpret factorial designs and explain how interactions between
    factors arise
-   Describe how matching and adjustment reduce confounding in
    observational studies
-   Explain how sample size is chosen to achieve desired precision or
    statistical power

# Experiments

## Observations Are Not Enough

::::: columns
::: {.column width="50%"}
-   [Observational studies]{.keyword}
    -   Reveal patterns in real-world data
    -   Researchers observe and measure variables as they naturally
        occur
-   But patterns alone do not tell us why they occur
-   [Association]{.keyword}: two variables change together
-   [Causation]{.keyword}: a change in one variable directly produces a
    change in another
-   Multiple explanations can produce the same association
-   Association does not imply causation
:::

::: {.column width="50%"}
![Correlation. Image: XKCD (CC BY-NC
2.5)](https://imgs.xkcd.com/comics/correlation.png){#fig-correlation}
:::
:::::

## Why Causal Inference Is Hard

::::: columns
::: {.column width="50%"}
-   [Confounding]{.keyword}: a third variable influences both variables
-   [Directionality]{.keyword}: cause and effect can be reversed
-   These problems cannot be resolved with more data alone
-   We need a way to isolate the effect of a single factor → experiments
:::

::: {.column width="50%"}
![Confounding example: a third factor (summer conditions) increases both
ice cream consumption and drowning deaths, creating a misleading
association without a direct causal
link.](images/spurious-correlation.png){#fig-spurious-correlation
fig-alt="Triangle diagram showing “In the summer…” at the top with arrows pointing to “ice cream” and “drowning deaths.” A crossed-out arrow between ice cream and drowning deaths indicates no direct causal relationship."}
:::
:::::

## Example: When Observational Studies Mislead

::::: columns
::: {.column width="50%"}
-   Observational studies suggested that hormone replacement therapy
    (HRT) reduced heart disease risk in women (Stamfer et al. 1991 New
    England J Med)

-   Women taking HRT had lower rates of heart disease

-   Conclusion (at the time): HRT protects against heart disease
:::

::: {.column width="50%"}
```{r}
#| label: fig-table-relative-risk
#| fig-cap: Annotated table from Postmenopausal estrogen therapy and cardiovascular disease showing relative risk (RR) estimates for cardiovascular outcomes by hormone use. The magenta box highlights the RR column label, blue boxes mark the reference group (no hormone use, RR = 1.0), and orange boxes highlight reduced RR estimates among current hormone users in the observational data (Stamfer et al. 1991 New England J Med).
#| fig-alt: Black-and-white table of cardiovascular disease outcomes by hormone use with colored annotations. A magenta box surrounds the "RR (95% CI)" column header. Blue boxes highlight RR values of 1.0 for the no hormone use group. Orange boxes highlight lower RR values for current hormone users across outcomes, indicating an apparent protective association.
#| fig-width: 8
#| fig-height: 3.387879

library(ggplot2)
library(jpeg)
library(grid)

img <- readJPEG("images/table-relative-risk-of-cardiovascular-disease.jpg")

ggplot() +
  annotation_custom(rasterGrob(img, interpolate = TRUE),
                    xmin = 0, xmax = 1, ymin = 0, ymax = 1) +
  # geom_rect(aes(xmin = 0.4, xmax = 0.56, ymin = .5, ymax = 0.8),
  #           color = "red", fill = NA, linewidth = 1.2) +
  # annotate("segment", )
  annotate("rect", xmin = 0.04, xmax = .17, ymin = .13, ymax = 0.175,
           color = "#CC79A7", fill = NA, linewidth = 1) +
    annotate("rect", xmin = 0.28, xmax = .36, ymin = .625, ymax = 0.673,
           color = "#CC79A7", fill = NA, linewidth = 1) +
  annotate("rect", xmin = 0.023, xmax = .14, ymin = .558, ymax = 0.606,
           color = "#0072B2", fill = NA, linewidth = 1) +  
  annotate("rect", xmin = 0.30, xmax = .34, ymin = .558, ymax = 0.606,
           color = "#0072B2", fill = NA, linewidth = 1) +
    annotate("rect", xmin = 0.023, xmax = .17, ymin = .38, ymax = .552,
           color = "#E69F00", fill = NA, linewidth = 1) +
    annotate("rect", xmin = 0.265, xmax = .377, ymin = .415, ymax = 0.515,
           color = "#E69F00", fill = NA, linewidth = 1) +
  # annotate("text", x = 0.5, y = .5, label = "Example", size = 6,
  #          color = "red") +
  coord_fixed(
    ratio = dim(img)[1] / dim(img)[2], 
    xlim = c(0, 1), 
    ylim = c(0, 1), 
    expand = FALSE
  ) +
  theme_void()
```
:::
:::::

## What Was the Problem?

-   Women who chose HRT differed in important ways:
    -   Higher socioeconomic status
    -   Better access to healthcare
    -   Healthier lifestyles overall
-   These differences (confounders) also reduce heart disease risk

## What Happened in an Experiment?

::::: columns
::: {.column width="50%"}
-   A randomized trial (Women’s Health Initiative) assigned HRT randomly

-   Result: HRT did **not** reduce heart disease risk (and increased
    some risks)

-   The original association was due to confounding, not causation
:::

::: {.column width="50%"}
![Kaplan–Meier Estimates of Cumulative Hazard Rates of CHD (Manson et
al. 2003 New England J Med) showing similar rate of chronic hearth
disease (CHD) among control and treatment
groups.](images/clipboard-2143964736.png){#fig-kaplan-meier-chd
fig-align="center" width="707"}
:::
:::::

## What Is an [Experiment]{.keyword}?

-   A study where researchers **actively impose conditions** on a system
-   Researchers **assign different conditions (treatments)** to
    experimental units
-   Outcomes are then compared across those conditions
-   This allows us to isolate the effect of a specific factor

## Why Experiments Work

-   Experiments are designed to support **causal inference**
-   By controlling how conditions are assigned, we reduce alternative
    explanations
-   Properly designed experiments break the link between confounders and
    treatment
-   Differences in outcomes can be attributed to the treatment

## Key Idea

-   Observational studies measure what already exists
-   Experiments **create conditions for comparison**
-   This is what allows us to move from association → causation

# Eliminating Bias

## Be Wary of Bias in Your Design

::::: columns
::: {.column width="50%"}
-   Biased experiments produce biased conclusions
-   They tell you about your design, not the real world
-   Bias must be addressed **before data are collected**
    -   It cannot be fixed afterward
-   Large $n$ does not solve bias
    -   It can make biased results more convincing
:::

::: {.column width="50%"}
![Comparison of sampling error and systematic error: each point
represents a sample estimate, and the center of the target represents
the true population value. Imprecision produces a wide spread of sample
estimates centered on the truth, whereas bias produces tightly clustered
estimates that are consistently offset from the
truth.](images/bias.png){#fig-bias
fig-alt="Two target diagrams side by side. In both, individual points represent sample estimates and the bullseye represents the true population value. The first target shows points widely scattered around the center, indicating sampling error or imprecision. The second shows a tight cluster of points located away from the center, indicating systematic error or bias."
fig-align="center"}
:::
:::::

## Design Features That Reduce Bias

-   Three core strategies:
    1.  [Controls]{.keyword}
        -   Provide a baseline for comparison
    2.  [Random assignment]{.keyword}
        -   Breaks the link between confounders and treatment
    3.  [Blinding]{.keyword}
        -   Prevents subjects and researchers from influencing outcomes

## Eliminating Bias: Strategy 1 — Use [Controls]{.keyword}

::::: columns
::: {.column width="50%"}
-   A control group provides a **baseline for comparison**\
-   Control units are treated as similarly as possible to treatment
    units
    -   Except for the treatment itself\
-   This allows us to isolate the effect of the treatment\
-   Without a control, we cannot determine cause and effect\
:::

::: {.column width="50%"}
:::
:::::

## What Makes a Good Control?

-   A good control matches the treatment group in all relevant ways\
-   The only systematic difference should be the treatment\
-   Poor controls introduce new differences (new confounding)\
-   Doing nothing is not always an appropriate control

## Control Example: [Placebo]{.keyword}

::::: columns
::: {.column width="50%"}
-   Outcomes can change simply because a treatment is given

    A placebo mimics the treatment without an active ingredient

    Good placebo: indistinguishable from the treatment

    Bad placebo: differs in noticeable ways (e.g., taste, side effects)
:::

::: {.column width="50%"}
![Illustration of the placebo effect: a participant shows a positive
response whether receiving an active drug or a placebo, demonstrating
how expectations can produce similar perceived and reported outcomes in
both groups.](images/placebo.png){#fig-placebo
fig-alt="Side-by-side realistic images of the same man smiling while being given a pill. In one panel he receives a plain white pill (placebo), and in the other a colored capsule (drug). His expression is similarly positive in both, indicating a comparable response regardless of treatment."}
:::
:::::

## Why Controls? Example: Independent Recovery

-   People often seek treatment at their worst.

-   Therefore, people often see their doctor when they are on their way
    to recovery.

-   To measure the effects of a new therapy, we need a comparable
    control group.

## Eliminating Bias: Strategy 2 — [Random Assignment]{.keyword}

::::: columns
::: {.column width="50%"}
-   Random assignment: assign treatments to units **by chance**
-   This breaks the link between confounders and treatment
-   Known and unknown differences are balanced **on average** across
    groups
-   Prevents systematic differences between groups
:::

::: {.column width="50%"}
:::
:::::

## Why Random Assignment Matters

-   Without randomization, group differences can reflect confounding
-   Non-random assignment can create bias
-   Example: assign treatment by last name
    -   May group family members or cultural backgrounds together
    -   Creates systematic differences between groups
-   Randomization prevents these patterns

## How to randomly assign

::::: columns
::: {.column width="50%"}
-   Identify the experimental units
-   Assign treatments using a random process
-   Goal: each unit has an equal chance of receiving each treatment
-   In practice:
    -   Use a random number generator
    -   Use software (e.g., R)
:::

::: {.column width="50%"}
``` r
library(tidyverse)
tibble(id = 1:10) |>                 # <1>
  mutate(                            # <2>
    treatment = sample(              # <3>
      x = c("control", "treatment"), # <4>
      size = n(),                    # <5>
      replace = TRUE                 # <6>
    )
  )
```

1.  Create a tibble (data frame) with one variable `id` with values of 1
    through 10
2.  Modify the variables in the tibble
3.  Add a new `treatment` variable using the `sample` function, which
    randomly draws values
4.  The values should consist of either `control` or `treatment`
5.  `n()` ensures the number of values in the sample matches the number
    of rows in the tibble
6.  `replace=TRUE` ensures true random assignment (which means it allows
    unequal counts)

![Illustration of the results of random
assignment.](images/clipboard-2471981232.png){#fig-random-assignment}
:::
:::::

## Results of random assignment

-   Randomization does not eliminate confounders
-   It **removes systematic bias** by breaking their association with
    treatment
-   Remaining differences are due to chance (sampling error)

## Eliminating Bias: Strategy 3 — [Blinding]{.keyword}

-   Blinding: keeping participants and/or researchers unaware of
    treatment assignment
-   Prevents expectations from influencing outcomes
    -   Participants may respond differently if they know their
        treatment
    -   Researchers may treat or measure subjects differently
-   Reduces bias introduced during data collection

## Types of Blinding

::::: columns
::: {.column width="50%"}
-   [Single-blind]{.keyword}: participants do not know their treatment
    -   Prevents subject expectations from influencing outcomes
-   [Double-blind]{.keyword}: neither participants nor researchers know
    -   Prevents both subject and researcher bias
-   Stronger blinding → less opportunity for bias\
:::

::: {.column width="50%"}
![Comparison of single-blind and double-blind study designs: in a
single-blind study, the participant is unaware of treatment assignment,
whereas in a double-blind study, both the participant and the researcher
are unaware.](images/single-double-blind.png){#fig-single-double-blind
fig-alt="A professional illustration showing a female researcher with black hair in a ponytail and a male participant seated at a table with pill and placebo bottles. Above, two labeled panels compare designs: under “Single Blind,” only the participant is blindfolded; under “Double Blind,” both the participant and researcher are blindfolded."
fig-align="center" width="510"}
:::
:::::

## Why Blinding Matters

::::: columns
::: {.column width="50%"}
-   Knowledge of treatment can influence:
    -   Behavior
    -   Reporting of symptoms
    -   Measurement of outcomes
-   Unblinded studies often show larger effects
    -   These may reflect bias, not true treatment effects
:::

::: {.column width="50%"}
![Illustration of lack of blinding: a participant believes they received
a placebo and attributes their symptoms to it, showing how expectations
can influence both perceived and reported
outcomes.](images/blinding-effect.png){#fig-blinding-effect
fig-alt="A realistic image of a woman with a frustrated expression, looking off to the side. A thought bubble above her head reads “No wonder I feel bad…” with a placebo symbol, indicating she believes she received a placebo and is interpreting and reporting her symptoms accordingly."}
:::
:::::

## Blinding in Practice

::::: columns
::: {.column width="50%"}
-   Blinding requires careful design

    -   Treatments must be indistinguishable

-   Placebos are often used to maintain blinding

-   If blinding fails, bias can re-enter the study
:::

::: {.column width="50%"}
![The limitations of blind trials are apparent when treatments are not
indistinguishable. Image: XKCD (CC BY-NC
2.5)](https://imgs.xkcd.com/comics/blind_trials.png){#fig-xkcd-blind-trials
fig-align="center" width="526"}
:::
:::::

# Sampling Error

## Reducing sampling error improves precision and power

-   Even unbiased experiments have variability among individuals
    (“noise”)
-   This variability creates sampling error in estimates
-   Sampling error reduces:
    -   Precision of estimates
    -   Power to detect treatment effects

## Holding conditions constant reduces noise but limits generality

-   Reduce noise by keeping conditions constant:
    -   Environment (e.g., temperature, humidity)
    -   Participant characteristics (e.g., age, sex, genotype)
-   Tradeoff:
    -   More control → less variability
    -   But results may not generalize broadly

## Overly narrow conditions can create bias in applicability

-   Restricting study populations limits who results apply to
-   Example:
    -   Many clinical trials historically included only men
    -   Results were applied broadly, including to women
-   Design decisions affect external validity

## Key design strategies reduce sampling error

-   Key design strategies:

    1.  Replication

    2.  Balance

    3.  Blocking

    4.  Using extreme treatments

-   Goal: reduce noise without sacrificing generality

## Replication is essential to separate signal from noise

-   [Replication]{.keyword}: applying each treatment to multiple
    experimental units
-   Without replication:
    -   Cannot distinguish treatment effects from random variation
-   More replication:
    -   More information
    -   Better estimates
    -   Higher power to detect real effects

## Replication depends on independent experimental units

-   Replicates must be **independent units**
-   Experimental unit:
    -   The unit assigned a treatment independently
-   Examples:
    -   Individual organism (if assigned independently)
    -   Group units: plot, cage, household, petri dish
-   Key rule:
    -   Individuals within the same unit are **not independent**

## Replication is not just "more individuals"

-   Multiple organisms ≠ multiple replicates
-   If organisms share the same environment:
    -   They are more similar to each other
    -   They count as **one replicate**
-   Must identify the correct experimental unit:
    -   Critical for design and analysis

## Example: Which designs are truly replicated?

::::: columns
::: {.column width="50%"}
-   Two growth chambers:
    -   Control vs Treatment (different light)
-   Multiple plants per chamber
    -   Share the same environment
-   Experimental unit = **chamber**, not plant
-   One chamber per treatment → **no replication**
-   Lighting differs between chambers
    -   Cannot separate treatment from chamber effect
:::

::: {.column width="50%"}
![Two growth chambers comparing control and treatment conditions, each
containing multiple Brassica rapa plants. Because all plants within a
chamber share the same environment—and the chambers differ in light
intensity—the chamber, not the plant, is the experimental unit. With
only one chamber per treatment, this design is
unreplicated.](images/growth-chambers.png){#fig-growth-chambers
fig-alt="Illustration of three experimental setups using potted plants with two fertilizer treatments shown by different pot colors. The top row shows one plant per treatment (no replication). The middle row shows multiple plants per treatment grouped into separate chambers (not independent, still unreplicated). The bottom row shows individual plants randomly assigned to treatments and interspersed, representing proper replication with independent experimental units."
fig-align="center"}
:::
:::::

## [Interspersion]{.keyword} signals proper replication

::::: columns
::: {.column width="50%"}
-   Proper replication shows **interspersion**:
    -   Treatments mixed across units
    -   Result of random assignment
-   Lack of interspersion:
    -   Warning sign of design problems
    -   Likely non-independence
:::

::: {.column width="50%"}
![Examples of good and poor experimental designs illustrating
interspersion of treatments. Top rows show properly interspersed designs
(completely randomized, randomized block, and systematic), while bottom
rows show problematic designs where treatments are segregated, clumped,
isolated, interdependent, or lack replication. Figure from Hurlbert
(1984).](images/interspersing.png){#fig-interspersing
fig-alt="Diagram comparing experimental layouts using black and white squares to represent two treatments. The top section labeled “Good design” shows treatments evenly mixed across units (completely randomized, randomized block, and systematic). The bottom section labeled “Poor design” shows treatments grouped or separated, including simple segregation, clumped segregation, isolation in separate chambers, interdependent replicates, and no replication."
fig-align="center" width="80%"}
:::
:::::

## [Pseudoreplication]{.keyword} leads to false precision

-   Treating non-independent units as independent =
    **pseudoreplication**
-   Examples:
    -   Treating plants within a chamber as separate replicates
    -   Repeated measurements on same individual
-   Consequences:
    -   Standard errors too small
    -   Overconfidence in results

## Why replication reduces standard error

$$
\operatorname{SE}_{\bar{Y}_1 - \bar{Y}_2} = \sqrt{s_p^2 \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}
$$

-   $n_1$, $n_2$ = number of independent replicates per treatment
-   Increasing sample size ↓ standard error
-   Lower standard error → clearer detection of differences

## Replication has practical limits

-   Increasing sample size improves inference
-   But comes with costs:
    -   Time
    -   Money
    -   Ethical considerations (e.g., animal use)
-   Goal:
    -   Sufficient replication to detect meaningful effects

## [Balanced]{.keyword} designs minimize sampling error

::::: columns
::: {.column width="50%"}
-   Balanced design = equal sample size in each treatment

-   Unbalanced design = unequal sample sizes

-   For a fixed total sample size:

    -   Standard error is **smallest when group sizes are equal**
    -   Balance optimizes precision of comparisons
:::

::: {.column width="50%"}
```{r}
#| label: fig-balanced-design
#| fig-cap: Balanced and unbalanced experimental designs illustrating allocation of sample size across treatments. In the balanced design, control and treatment groups have equal numbers of independent experimental units (n = 6 each). In the unbalanced design, most units are assigned to the control group (n = 10) and few to the treatment group (n = 2). Circles represent control units and squares represent treatment units.
#| fig-alt: Diagram comparing balanced and unbalanced experimental designs. The balanced design shows equal numbers of units in control (circles) and treatment (squares). The unbalanced design shows many control units and few treatment units, illustrating unequal sample sizes across treatments.
#| message: false

library(ggplot2)
library(dplyr)
library(tibble)

#-----------------------------
# DATA
#-----------------------------

# Balanced: 6 vs 6
balanced <- tibble(
  panel = "Balanced design",
  x = c(1:6, 8:13),
  treatment = c(rep("Control", 6), rep("Treatment", 6))
)

# Unbalanced: 10 vs 2
unbalanced <- tibble(
  panel = "Unbalanced design",
  x = c(1:10, 12:13),
  treatment = c(rep("Control", 10), rep("Treatment", 2))
)

df <- bind_rows(balanced, unbalanced)

labels <- df |> 
  summarize(
    x = mean(x),
    label = paste0("n = ", n()),
    .by = c(panel, treatment)
  )

ggplot(df, aes(x = x, y = 1)) +
  geom_point(
    aes(shape = treatment, fill = treatment),
    size = 10, color = "black", stroke = 1
  ) +
  scale_shape_manual(values = c(21, 22), guide = "none") +
  scale_fill_manual(values = c("#56B4E9", "#E69F00"), guide = "none") +
  facet_wrap(~ panel, ncol = 1) +
  geom_text(  # treatment labels
    data = labels,
    aes(label = treatment),
    size = 9, position = position_nudge(y = -.5)
  ) +
  geom_text(  # n labels
    data = labels,
    aes(label = label),
    size = 9, position = position_nudge(y = -1)
  ) +
  scale_x_continuous(expand = expansion(c(.1,.2),0)) +
  scale_y_continuous(expand = expansion(c(.4,0.8),0)) +
  theme_void() +
  theme(
    strip.text = element_text(size = 30, face = "bold"),
    plot.margin = margin(t = 30),
    panel.spacing = unit(2, "lines"),
    strip.clip = "off"
  )


```
:::
:::::

## Why balance improves precision

::::: columns
::: {.column width="50%"}
$$
\operatorname{SE}_{\bar{Y}_1 - \bar{Y}_2} = \sqrt{s_p^2 \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}
$$

-   For fixed $n_1 + n_2$:
    -   SE minimized when $n_1 = n_2$
-   Example (total $n = 20$):
    -   Balanced: $n_1=10$, $n_2=10$ → smaller SE
    -   Unbalanced: $n_1=19$, $n_2=1$ → much larger SE
:::

::: {.column width="50%"}
-   Estimating a difference requires:
    -   Precise estimate of **both** means
-   Unbalanced design:
    -   One group well estimated
    -   Other poorly estimated → weak comparison
-   Balance allocates effort efficiently
:::
:::::

## Balance is **optimal** but not strictly required

-   Increasing sample size improves precision:
    -   Even if added to only one group
-   But for a fixed total sample size:
    -   Equal allocation is **optimal**
-   Additional benefit:
    -   Statistical methods are more robust
    -   Especially when variances differ between groups

## Blocking reduces noise from known sources of variation

-   [Blocking]{.keyword}: group similar experimental units into
    **blocks**

-   Units within a block:

    -   Share location or other characteristics
    -   Are more similar to each other than to units in other blocks

-   Goal:

    -   Remove variation not caused by the treatment
    -   Increase precision and power

## How blocking works

::::: columns
::: {.column width="50%"}
-   Within each block:

    -   Assign treatments **randomly**
    -   Treatments are interspersed within the block

-   Analyze differences **within blocks**, not across all units

-   Conceptually:

    -   Repeat the same experiment in each block
    -   Compare treatments under similar conditions
:::

::: {.column width="50%"}
![Comparison of experimental designs with and without blocking. Blocking
groups similar individuals before random assignment, allowing treatment
comparisons within blocks and reducing variation among groups. Image:
JHK111 (CC0 1.0 Universal)](images/blocking.jpg){#fig-blocking
fig-alt="Diagram comparing no blocking and blocking designs. In the no blocking design, all individuals are randomly divided into treatment and control. In the blocking design, individuals are first grouped into blocks, then treatments are randomly assigned within each block."}
:::
:::::

## When is blocking useful?

:::::: columns
::: {.column width="40%"}
-   Use blocking when:
    -   Units differ due to known factors (e.g., location, time, group)
-   Examples of blocks:
    -   Field plots in the same area
    -   Animals from the same litter
    -   Patients from the same clinic
    -   Experiments run on the same day
-   Key condition:
    -   Units within blocks are similar
    -   Blocks differ from each other
:::

:::: {.column width="60%"}
::: {#fig-elephants layout-ncol="2"}
![Without blocking: diet pills vs placebo on weight loss. Individuals
are not grouped, so variation among individuals obscures differences
between treatments. Image: JHK111 (CC0 1.0
Universal).](images/without-blocking.jpg){#fig-without-blocking
fig-alt="Scatterplot of weight loss by individuals without blocking. Blue points represent placebo and green points represent diet pills, with substantial overlap between groups making treatment differences difficult to distinguish."
fig-align="center" width="475"}

![With blocking: diet pills vs placebo on weight loss. Individuals are
grouped into blocks (females and males), and treatment differences are
clearer within each block. Image: JHK111 (CC0 1.0
Universal).](images/with-blocking.jpg){#fig-with-blocking
fig-alt="Scatterplot of weight loss grouped into two blocks labeled females and males. Within each block, blue points represent placebo and green points represent diet pills, showing clearer separation between treatments compared to the unblocked design."
fig-align="center" width="434"}

Comparison of experimental designs with and without blocking. Without
blocking, variation among individuals obscures treatment differences.
With blocking, individuals are grouped into more similar subsets, making
treatment effects easier to detect within each block.
:::
::::
::::::

## Example: Extreme treatments reveal nitrogen effects

::::: columns
::: {.column width="50%"}
-   Clark and Tilman (2008) studied whether nitrogen addition reduces
    plant diversity

-   Typical (background) N deposition: \~1–10 kg N ha⁻¹ yr⁻¹

-   Experimental treatments: Up to **100 kg N ha⁻¹ yr⁻¹** (extreme)

-   Why use extreme levels?

    -   Amplify the response
    -   Make treatment effects easier to detect

-   Result: Clear decline in species richness with higher N
:::

::: {.column width="50%"}
![Relationship between nitrogen input and plant species loss in
grassland ecosystems. Species loss increases with nitrogen addition,
with stronger effects at higher (extreme) nitrogen levels. Points
represent observations and lines show fitted trends. Adapted from Clark
and Tilman
(2008).](images/nitrogen-extreme-treatment.jpg){#fig-nitrogen-extreme-treatment
fig-alt="Scatterplot showing species loss (proportion) versus nitrogen input rate. Points are scattered across increasing nitrogen levels, with fitted lines indicating an upward trend in species loss as nitrogen input increases, especially at higher application rates."
fig-align="center"}
:::
:::::

## Extreme treatments make effects easier to detect

-   Treatment effects are easiest to detect when they are **large**

-   Small differences:

    -   Hard to distinguish from random variation
    -   Require larger sample sizes

-   Large differences:

    -   Stand out against noise
    -   Increase power to detect an effect

-   Strategy:

    -   Include **extreme treatment levels**

## Extreme treatments increase power, but with tradeoffs

-   Stronger treatments → larger response differences
    -   Higher probability of detecting an effect
-   Useful as a first step:
    -   Does this variable affect the response at all?
-   Caution:
    -   Effects may not scale linearly
    -   Extreme treatments may not reflect realistic conditions
-   Balance:
    -   Detection vs realism

# Experiments with More Than One Factor

## Experiments can include more than one factor

::::: columns
::: {.column width="50%"}
-   A [factor]{.keyword}:
    -   A single treatment variable of interest
-   Many experiments include **multiple factors**
    -   More efficient:
        -   Answer multiple questions at once
        -   Use time, materials, and effort more effectively
-   Example idea:
    -   Temperature + nutrients
    -   Light + water
:::

::: {.column width="50%"}
![Factorial design illustrating the combined effects of light and water
on plant growth. Each panel represents a different combination of low
and high levels of both factors, showing how growth depends on their
interaction.](images/factorial-example.png){#fig-factorial-example
fig-alt="A 2×2 grid showing plant growth under combinations of low and high light and water. Plants are smallest under low light and low water, larger with either factor increased, and largest when both light and water are high."}
:::
:::::

## Factorial designs test combinations of factors

::::: columns
::: {.column width="50%"}
-   [Factorial design]{.keyword}:
    -   Includes **all combinations** of treatment levels
-   Example structure (2 factors):
    -   Factor A: A₁, A₂
    -   Factor B: B₁, B₂
-   Treatments:
    -   A₁B₁, A₁B₂, A₂B₁, A₂B₂
-   Key advantage:
    -   Can test **interactions** between factors
:::

::: {.column width="50%"}
```{r}
#| label: fig-factorial-table
#| fig-cap: Factorial design with two variables (A and B), each with two levels. Rows represent levels of Variable A and columns represent levels of Variable B. Each cell shows a treatment combination (e.g., A₁B₁), representing one unique combination of factor levels included in the experiment.
#| fig-alt: Table illustrating a 2×2 factorial design. Rows are labeled A₁ and A₂ for Variable A, and columns are labeled B₁ and B₂ for Variable B. Each cell contains a treatment combination (A₁B₁, A₁B₂, A₂B₁, A₂B₂), showing all possible combinations of the two factors.

library(gt)
library(tibble)

factorial_tbl <- tibble(
  group = "Variable A",
  level_a = c("A₁", "A₂"),
  b1 = c("A₁B₁", "A₂B₁"),
  b2 = c("A₁B₂", "A₂B₂")
)

factorial_tbl |>
  gt(
    rowname_col = "level_a",
    groupname_col = "group"
  ) |>
  tab_header(
    title = "Factorial design: two factors with two levels each",
  ) |>
  tab_spanner(
    label = "Variable B",
    columns = c(b1, b2)
  ) |>
  cols_label(
    b1 = "B₁",
    b2 = "B₂"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(
      cells_row_groups(groups = "Variable A"),
      cells_column_spanners(spanners = "Variable B")
    )
  ) |>
  tab_options(
    table.font.size = px(28),
    column_labels.font.size = px(28),
    row_group.font.size = px(28),
    stub.font.size = px(28),
    data_row.padding = px(14),
    column_labels.padding = px(14),
    table.width = pct(70),
    container.padding.y = px(40)
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  opt_stylize(style = 6, color = "blue", add_row_striping = FALSE)


```
:::
:::::

## Interactions: when effects depend on each other

::::: columns
::: {.column width="50%"}
-   [Interaction]{.keyword}: Effect of one factor depends on another
    factor
-   Without interaction:
    -   Effects are independent and additive
-   With interaction:
    -   Combined effect differs from separate effects
-   Only detectable with **factorial design**
-   [Examples of types of interactions (Duda et al.
    2023)](https://www.nature.com/articles/s41598-023-47057-0/figures/1)
:::

::: {.column width="50%"}
```{r}
#| label: fig-interaction-example
#| fig-cap: Hypothetical interaction between Variable A and Variable B. The effect of Variable B on the response differs depending on the level of Variable A, as shown by the non-parallel lines.
#| fig-alt: Line graph showing response versus Variable B (B₁, B₂) with separate lines for Variable A (A₁, A₂). The line for A₁ increases from B₁ to B₂, while the line for A₂ decreases, illustrating a non-parallel interaction effect.

library(ggplot2)
library(tibble)

# Hypothetical interaction data
df <- tibble(
  A = factor(rep(c("A1", "A2"), each = 2), levels = c("A1", "A2")),
  B = factor(rep(c("B1", "B2"), times = 2), levels = c("B1", "B2")),
  response = c(5, 9,   # A₁: increases with B
               5, 2)   # A₂: decreases with B (interaction)
)

# Plot
ggplot(df, aes(x = B, y = response, group = A, color = A)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c("A1" = "#0072B2", "A2" = "#D55E00"),
    name = "Variable A"
  ) +
  labs(
    title = "Interaction between Variable A and Variable B",
    x = "Variable B",
    y = "Response"
  ) +
  theme_classic(base_size = 18)
```
:::
:::::

## Example: 4-factor factorial experiment (smoking reduction)

::::: columns
::: {.column width="50%"}
-   Study in Cook *et al.* (2015) *Addiction*

-   Outcome: % reduction in cigarettes/day

-   **4 factors (2 levels each: yes vs no):**

    -   Nicotine patch
    -   Nicotine gum
    -   Motivational interviewing
    -   Behavioral reduction

-   Design: **2 × 2 × 2 × 2 = 16 combinations**

-   Key idea: Effects depend on **combinations of treatments
    (interactions)**
:::

::: {.column width="50%"}
![Mean percent reduction in cigarettes per day for all combinations of
four treatments (nicotine patch, nicotine gum, motivational
interviewing, and behavioral reduction). Each panel represents a
different combination of patch and motivational interviewing, with bars
showing gum and behavioral reduction combinations. Results illustrate
how treatment effects vary across combinations, indicating interactions
among factors. Adapted from Cook et al.
(2015).](images/factorial-example-smoking.png){#fig-factorial-example-smoking
fig-alt="Multi-panel bar chart showing mean percent reduction in cigarettes per day across combinations of four treatments: nicotine patch, nicotine gum, motivational interviewing, and behavioral reduction. Each panel corresponds to patch and motivational interviewing conditions, with bars representing gum and behavioral reduction combinations. Differences among bars indicate that treatment effects depend on combinations of factors."
width="75%"}
:::
:::::

# What if You Can’t Do an Experiment?

## When experiments are not possible

-   Use [observational studies]{.keyword}
    -   Researcher does **not assign treatments**
    -   Subjects come as they are
-   Strengths:
    -   Detect real-world patterns
    -   Generate hypotheses
-   Limitation:
    -   Cannot use **randomization**
    -   → greater risk of bias

## Observational studies still use good design principles

-   Apply as many experimental design features as possible:
    -   Controls
    -   Blinding (when possible)
    -   Replication, balance, blocking
-   Key missing feature:
    -   **Randomization**
-   Biggest challenge:
    -   [Confounding variables]{.keyword}

## Strategy 1: Matching

::::: columns
::: {.column width="50%"}
-   [Matching]{.keyword}: Pair each treated individual with a similar
    control
-   Match on known confounders:
    -   Age, sex, weight, background, etc.
-   Common in: **Case–control studies**
-   Benefits:
    -   Reduces bias
    -   Reduces sampling error (like blocking)
-   Limitation:
    -   Only controls **known** confounders
:::

::: {.column width="50%"}
![Illustration of matching in a case–control study. Individuals are
paired so that cases and controls have the same characteristics (sex,
race, and age), reducing confounding. From Dey et al. (2020) Chest
J.](images/matching.jpg){#fig-matching
fig-alt="Grid of human icons arranged in pairs connected by arrows. Each pair matches individuals with the same sex (male or female), race (light or dark shading), and age (with or without a cane), illustrating one-to-one matching on multiple confounding variables."
fig-align="center" width="584"}
:::
:::::

## Strategy 2: Adjustment

::::: columns
::: {.column width="50%"}
-   [Adjustment]{.keyword}: Use statistical methods to control for
    confounders
-   Example approach:
    -   Compare groups at the **same value** of a confounder (e.g., age)
-   Methods:
    -   Regression
    -   Analysis of covariance (ANCOVA)
-   Key requirement:
    -   Groups must **overlap** in confounder values
:::

::: {.column width="50%"}
```{r}
#| label: fig-adjustment-penguins
#| fig-cap: Relationship between flipper length and body mass in Adelie penguins, with separate regression lines for males and females. Points show individual observations, and lines show fitted values from a model including sex and flipper length. The vertical line marks a common flipper length, and highlighted points show predicted (adjusted) body mass for each sex at that value, illustrating comparison after accounting for flipper length.
#| fig-alt: Scatterplot of body mass versus flipper length for Adelie penguins, colored by sex, with separate regression lines for males and females. A dashed vertical line marks a common flipper length, and points on each line at that position indicate predicted body mass for each sex, showing adjusted comparison controlling for flipper length.
#| fig-height: 5
  
library(tidyverse)
library(palmerpenguins)
library(see)

# Data: Adelie only, remove missing values
df <- penguins |>
  filter(species == "Adelie") |>
  drop_na(flipper_length_mm, body_mass_g, sex)

# Fit ANCOVA model
mod <- lm(body_mass_g ~ sex + flipper_length_mm, data = df)

# Choose a common flipper length (approximate center of data)
x_ref <- 190

# Prediction data
pred_df <- tibble(
  sex = levels(df$sex),
  flipper_length_mm = x_ref
)

# Add predicted values
pred_df$body_mass_g <- predict(mod, newdata = pred_df)

# Plot
ggplot(df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) +
  geom_point(alpha = 0.6) +
  
  # Regression lines from ANCOVA model
  geom_smooth(method = "lm", se = FALSE) +
  
  # Vertical reference line
  geom_vline(xintercept = x_ref, linetype = "dashed") +
  
  # Predicted means at common flipper length
  geom_point(
    data = pred_df,
    aes(x = flipper_length_mm, y = body_mass_g, color = sex),
    size = 4
  ) +
  
  labs(
    title = "ANCOVA: Adjusting for flipper length when\ncomparing sexes",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Sex"
  ) +
  scale_color_oi() +
  theme_minimal(base_size = 18)

```
:::
:::::

## Limits of observational studies

-   Observational studies can reveal important patterns
-   But without randomization:
    -   Confounding cannot be fully eliminated
-   Strongest inference:
    -   Experiments \> observational studies
-   Best use:
    -   Identify relationships
    -   Generate hypotheses for experiments

# Choosing a Sample Size

## Choosing a sample size matters

-   Goal: Choose enough samples to get **useful results**
-   Too small:
    -   Cannot detect effects
    -   Very wide confidence intervals
-   Too large:
    -   Wastes time, money, and resources
    -   May raise ethical concerns
-   Key question: How many replicates per treatment?

## Two ways to plan sample size

-   Plan for [precision]{.keyword}
    -   Want a **narrow confidence interval**
-   Plan for [power]{.keyword}
    -   Want a high probability of detecting a real effect
-   Focus here:
    -   Comparing **two means**

## Planning for precision

-   Goal: Estimate the difference in means:

$$
\mu_1 - \mu_2
$$

-   Use sample estimate:

$$
\bar{Y}_1 - \bar{Y}_2
$$

-   Want a **95% confidence interval** with small width

## Margin of error drives sample size

::::: columns
::: {.column width="50%"}

-   Confidence interval form:

$$
(\bar{Y}_1 - \bar{Y}_2) \pm \text{margin of error}
$$

-   Margin of error ≈

$$
2 \times \operatorname{SE}
$$
:::

::: {.column width="50%"}

-   Standard error:

$$
\operatorname{SE} = \sqrt{\frac{2\sigma^2}{n}}
$$

-   Larger $n$ → smaller SE → narrower interval
:::
:::::






## Sample size formula (approximate)

-   Solve for sample size per group:

$$
n \approx \frac{8\sigma^2}{(\text{margin of error})^2}
$$

-   Interpretation:
    -   Larger variability ($\sigma$) → need larger $n$
    -   Higher precision (smaller margin) → need larger $n$

## Practical challenge

-   $\sigma$ is **unknown**
    -   Use:
        -   Pilot studies
        -   Previous research
        -   Educated guess
-   Result:
    -   Sample size planning is approximate

## Choosing sample size for power

::::: columns
::: {.column width="50%"}
-   Goal: Choose $n$ so you can **detect a meaningful effect**
-   You must specify:
    -   **Effect size** (what difference matters biologically)
    -   **Variability** ($\sigma$; from pilot data or past studies)
    -   **Significance level** ($\alpha$, usually 0.05)
    -   **Desired power** (commonly **80%**)
        -   80% chance of rejecting a false null
        -   20% chance of missing a real effect (Type II error)
:::

::: {.column width="50%"}
-   How you do it:
    -   Use software (e.g., R, online calculators)
    -   Input these values → solve for required $n$
-   Key idea:
    -   Larger effect → smaller $n$
    -   Higher variability → larger $n$
    -   Higher power → larger $n$
:::
:::::

## More data improves precision

::::: columns
::: {.column width="50%"}
-   Small sample size:
    -   Very wide confidence intervals
-   Increasing $n$:
    -   Rapid improvement at first
-   Large $n$:
    -   Diminishing returns
:::

::: {.column width="50%"}
```{r}
#| label: fig-sample-size-precision
#| fig-cap: Relationship between sample size per treatment and expected precision of a 95% confidence interval for the difference in means. Precision is expressed as the margin of error divided by σ. Precision improves rapidly at small sample sizes and then more slowly, illustrating diminishing returns.
#| fig-alt: Line graph showing expected margin of error divided by sigma versus sample size per treatment. The curve decreases steeply from small sample sizes and then levels off, indicating that increases in sample size lead to diminishing improvements in precision.

library(tidyverse)

# Sequence of sample sizes per group
df <- tibble(
  n = 2:20
) |>
  mutate(
    df = 2 * n - 2,
    t_crit = qt(0.975, df = df),
    moe_over_sigma = t_crit * sqrt(2 / n)
  )

ggplot(df, aes(x = n, y = moe_over_sigma)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.8, color = "#D9481C") +
  scale_x_continuous(
    breaks = c(2, 5, 10, 15, 20),
  ) +
  scale_y_continuous(
    breaks = 0:4,
    limits = c(0, NA),
    expand = expansion(c(0,.1))
  ) +
  labs(
    x = expression("Sample size " * n * " in each treatment"),
    y = expression("Expected margin of error / " * sigma),
    title = "Precision improves with sample size",
  ) +
  theme_minimal(base_size = 22) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.line = element_line(),
    axis.ticks = element_line(),
    axis.ticks.length = unit(0.25, "cm"),
    axis.title.x = element_text(face = "bold", margin = margin(t = 16)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 16)),
    plot.margin = margin(10, 20, 10, 10)
  )

```
:::
:::::

## Diminishing returns

-   Precision improves quickly at first:
    -   e.g., $n = 2 \rightarrow 5$
-   Then slows:
    -   e.g., $n = 15 \rightarrow 20$
-   Each additional replicate adds **less new information**
-   Tradeoff:
    -   Precision vs cost

# Summary of Considerations in Experimental Design

## Summary: Designing studies and making inference

::::: columns
::: {.column width="50%"}
-   Experiments assign treatments and enable causal inference

-   Bias is reduced through controls, randomization, and blinding

-   Randomization balances confounding variables on average

-   Observational studies lack randomization and have weaker inference

-   Confounding in observational studies is reduced by matching and
    adjustment
:::

::: {.column width="50%"}
-   Sampling error is reduced by replication, balance, and blocking

-   Extreme treatments increase the ability to detect effects

-   Factorial designs test multiple factors and their interactions

-   Sample size is planned for precision or power

-   Study design involves tradeoffs among precision, cost, and
    feasibility
:::
:::::
