---
title: "Lab 4: Graphing with ggplot"
description: "An introduction to creating scatterplots in R using ggplot2 and a layered approach to graphing."
format: html
execute:
  echo: true
---

## Learning Objectives

By the end of this lab, students will be able to:

-   Describe the basic idea of building graphs in layers using ggplot2.
-   Create a ggplot object by specifying a dataset and aesthetic mappings.
-   Map numerical variables to the x- and y-axes to produce a scatterplot.
-   Map a categorical variable to color to represent groups in the data.
-   Use `facet_wrap()` to create small-multiple scatterplots.
-   Recreate a scatterplot with aesthetics and facets using a new dataset.

## Getting started

Before you begin the lab activities, complete the following steps to make sure you are fully set up.

1.  **Get the Lab Handout.**

    Pick up a physical copy of the lab worksheet, or print one if you are working outside of class.\
    <a href="lab-04-worksheet.pdf" download class="btn btn-outline-primary mt-3">Download Lab Worksheet (PDF, if needed)</a>

2.  **Open Posit Cloud and Start Lab 4.**

    Log in to **Posit Cloud** and navigate to the **course workspace**, and start the Lab 4 assignment.

3.  **Install required packages**

    Install the required packages for this lab, including **tidyverse** and **gapminder** using the Packages pane or by running the following code ***in the Console***:

    ```{r}
    #| label: install-packages
    #| eval: false
    #| echo: true

    install.packages("tidyverse", "gapminder")
    ```

4.  **Create an R Script**

    Create a new R script and save it as **lab-3-script.R**.

5.  **Load the packages**

    Copy and paste this code to your R script and run it to load the packages. Remember to run both library commands (Ctrl+Enter on one line, then the next).

    ```{r}
    #| label: load-packages
    #| echo: true

    # load packages --------------------------------------------------------------

    library(tidyverse)
    library(gapminder)
    ```

::: callout-note
## Checkpoint

At this point, you should have:

-   These instructions open in a web browser.
-   Your Lab 4 project open in Posit Cloud in another browser window.
-   The required packages installed and loaded without errors
-   The Lab 4 worksheet in front of you.

Do not continue until all of the above steps are working correctly.
:::

## Overview: From Data to Graphs

## The Gapminder Dataset

## How ggplot Builds a Graph

## Starting a Plot with `ggplot()`

## Mapping Variables with `aes()`

## Adding Points with `geom_point()`

## Using Color to Represent Groups

## Creating Small Multiples with `facet_wrap()`

## Interpreting the Code–Graph Connection

## Guided Practice: Modifying an Existing Plot

## Independent Practice: A New Scatterplot

## Key Takeaways

## Submission and Wrap-up
