---
title: |
  | Lecture 14
  | Inference for Population Means: 
  | Confidence Intervals and t-Tests
subtitle: "ABD 3e Chapter 11"
description: "Introduces statistical inference for normally distributed data, including confidence intervals for the mean and hypothesis tests (one-sample, independent two-sample, and paired t-tests) used to evaluate differences in population means."
format:
  revealjs:
    fig-width: 8
    fig-height: 6
editor: 
  markdown: 
    wrap: 72
---

## Learning Objectives

By the end of this lecture, you should be able to:

-   Explain why the **t-distribution** is used instead of the normal
    distribution when the population standard deviation is unknown
-   Describe how **degrees of freedom** influence the shape of the
    t-distribution
-   Construct and interpret a **confidence interval for a population
    mean**
-   State null and alternative hypotheses for a **one-sample t-test**
-   Interpret the results of a t-test using **t, degrees of freedom,
    p-value, and confidence interval**

## From Probability to Statistical Inference

::::: columns
::: {.column width="50%"}
-   Previous lectures introduced probability distributions and
    hypothesis testing for categorical data and proportions.
-   Today we apply the same inferential framework to **means of
    numerical variables**.
-   The goal is to estimate and test hypotheses about a population mean
    using sample data.
:::

::: {.column width="50%"}
![Conceptual diagram showing how a population mean (parameter) and a
sample mean (statistic) are obtained from a population and a sample.
Source: Illustration generated by
ChatGPT.](images/population-sample-statistic.png){#fig-sample-to-population-inference
fig-alt="Diagram illustrating statistical inference. A large circle in the upper left contains many colored dots representing a population. An arrow labeled “Sample” points to a smaller square in the upper right containing a subset of dots representing a sample. A downward arrow labeled “Parameter” leads from the population to a graph labeled “Population Parameter,” which shows a single point marking the population mean. Another downward arrow labeled “Estimation” leads from the sample to a graph labeled “Sample Statistic,” which shows a point with a vertical line range indicating uncertainty in the estimated mean. Both graphs have a vertical axis labeled “Mean (μ)” and no horizontal axis."
fig-align="center"}
:::
:::::

## Sampling Distributions (Review)

::::: columns
::: {.column width="50%"}
-   Statistics such as the sample mean vary from sample to sample due to
    sampling variability.
-   The **sampling distribution of the mean** describes the distribution
    of possible sample means.
-   If the population is normal, the sampling distribution of the mean
    is also normal.
:::

::: {.column width="50%"}
![Sampling distribution of the sample mean. Repeated random samples from
the same population produce different sample means; the distribution of
those means forms the sampling distribution. *Source: GeeksforGeeks,
"[Sampling
Distribution](https://www.geeksforgeeks.org/maths/sampling-distribution/)"*.](images/sampling-distribution.png){#fig-sampling-distribution
fig-alt="Diagram showing a population of 5th-grade students, many random samples drawn from that population, the sample mean from each sample, and a histogram of those sample means, illustrating the sampling distribution of the mean."
width="1400"}
:::
:::::

## Standard Error of the Mean (Review)

::::: columns
::: {.column width="50%"}
-   The spread of the sampling distribution is measured by the [standard
    error]{.keyword} (SE) of the mean.
-   Standard error quantifies how far sample means tend to vary from the
    population mean.
-   Larger sample sizes reduce the standard error and improve precision
    of the estimate.
:::

::: {.column width="50%"}
```{r}
#| label: fig-sampling-distribution-by-n
#| fig-height: 7.5
#| fig-width: 7.5
#| fig-cap: "Sampling distributions of the mean penguin body mass for two sample sizes. Smaller samples (n = 10) produce more variable sample means than larger samples (n = 100)."

library(dplyr)
library(ggplot2)
library(palmerpenguins)

set.seed(123)

n_reps <- 1000
sample_sizes <- c(10, 100)

sampling_means <- bind_rows(
  lapply(sample_sizes, function(n) {
    tibble(rep = 1:n_reps) |>
      rowwise() |>
      mutate(
        mean_mass = penguins |>
          filter(!is.na(body_mass_g)) |>
          slice_sample(n = n) |>
          summarize(mean(body_mass_g)) |>
          pull(),
        n = paste("n =", n)
      ) |>
      ungroup()
  })
)

# Compute per-facet x-range and max bin count (approx) using ggplot2's binning
bin_width <- (max(sampling_means$mean_mass) - min(sampling_means$mean_mass)) / 30

# Create histogram bin counts per facet
hist_counts <- sampling_means |>
  mutate(
    bin = floor((mean_mass - min(mean_mass)) / bin_width)
  ) |>
  group_by(n, bin) |>
  summarize(count = n(), .groups = "drop")

arrow_df <- hist_counts |>
  group_by(n) |>
  summarize(
    y = 0.20 * max(count),
    x_min = min(sampling_means$mean_mass[sampling_means$n == unique(n)]),
    x_max = max(sampling_means$mean_mass[sampling_means$n == unique(n)]),
    .groups = "drop"
  )

ggplot(sampling_means, aes(x = mean_mass)) +
  geom_histogram(
    bins = 30,
    fill = "#0072B2",
    color = "white"
  ) +
  geom_segment(
    data = arrow_df,
    aes(x = x_min, xend = x_max, y = y, yend = y),
    color = "#D55E00",
    linewidth = 2,
    arrow = arrow(
      ends = "both", type = "closed", length = unit(0.2, "inches")
    )
  ) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(
    title = "Effect of Sample Size on the Sampling\nDistribution of the Mean",
    x = "Sample mean body mass (g)",
    y = "Count"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    strip.text = element_text(size = rel(1.35))
  )

```
:::
:::::

## When the Population SD Is Unknown

::::: columns
::: {.column width="50%"}
### **Population standard error**

-   The standard error of the mean (SE) describes the expected
    variability of sample means around the population mean.

-   If the population standard deviation (SD) is known, the population
    SE is calculated using $\sigma$.

-   This formula describes the theoretical variability of the sample
    mean when the population variability is known.
:::

::: {.column width="50%"}
### **Estimated standard error**

-   In practice, the population standard deviation $\sigma$ is almost
    never known.

-   Instead, we estimate variability using the sample standard deviation
    $s$ calculated from the observed data.

-   Replacing $\sigma$ with $s$ gives an estimated standard error of the
    mean.
:::
:::::

::::: columns
::: {.column width="50%"}
$$
SE = \frac{\sigma}{\sqrt{n}}
$$
:::

::: {.column width="50%"}
$$
SE = \frac{s}{\sqrt{n}}
$$
:::
:::::

## Standardizing the Mean (Review: Z statistic)

::::: columns
::: {.column width="60%"}
-   In statistics, we often want to test:

    > Does a sample provide evidence that a population mean differs from
    > a hypothesized value.

-   To evaluate how unusual a sample mean is, we measure its distance
    from the pop. mean **in units of standard error**.

-   If the population standard deviation $\sigma$ were known, we could
    use the **Z statistic**.

$$
Z = \frac{\bar{x} - \mu}{\sigma / \sqrt{n}}
$$

-   The Z statistic measures how many standard errors the sample mean is
    from the population mean.
:::

::: {.column width="40%"}
The Z distribution

```{r}
#| label: fig-z-distribution
#| fig-cap: Standard normal (Z) distribution used to measure how far a sample mean is from a hypothesized population mean in standard error units.
#| fig-alt: Bell-shaped standard normal distribution centered at zero, with the horizontal axis labeled Z and the vertical axis labeled probability density.

library(ggplot2)
library(dplyr)

# Generate Z values
x <- seq(-3.2, 3.2, length.out = 1000)

# Degrees of freedom for t distribution
df_t <- 5

# Create stacked data frame
df <-
  bind_rows(
    z = tibble(x = x, density = dnorm(x)),
    t = tibble(x = x, density = dt(x, df = df_t)),
    .id = "distribution"
  ) |> 
  mutate(distribution = forcats::as_factor(distribution))

p <-
  df |> 
  ggplot(aes(x = x, y = density, color = distribution)) +
  scale_x_continuous(
    breaks = -3:3,
    expand = 0
  ) +
  scale_y_continuous(
    breaks = seq(0, 0.4, by = 0.1),
    limits = c(0, 0.4),
    expand = 0
  ) +
  ggthemes::scale_color_colorblind(drop = FALSE) +
  guides(color = "none") +
  labs(y = "Probability density") +
  theme_classic(base_size = 18) +
  theme(
    axis.line = element_line(linewidth = 1),
    axis.ticks = element_line(linewidth = 1),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16)
  )

p +
  geom_line(
    data = ~ filter(.x, distribution == "z"),
    linewidth = 1.5
  ) +
  labs(x = "Z")
```
:::
:::::

## Z Scores vs Z Statistics

-   The previous equation may look unfamiliar.

-   In earlier lectures, we standardized **individual observations**
    using the population standard deviation $\sigma$.

-   When working with **sample means**, we instead standardize using the
    **standard error of the mean**.

-   Sample means vary less than individual observations because they are
    averages of multiple values.

-   The variability of sample means decreases as sample size increases,
    which is why the denominator includes $\sqrt{n}$.

::: {style="margin: 1em 2em;"}
| Situation | Formula | Term | Interpretation |
|------------------|------------------|------------------|-------------------|
| Individual observation | $Z = \frac{x - \mu}{\sigma}$ | Z score | Distance of an observation from the population mean in **standard deviations** |
| Sample mean | $Z = \frac{\bar{x} - \mu}{\sigma/\sqrt{n}}$ | Z statistic | Distance of the sample mean from the population mean in **standard errors** |
:::

## Replacing $\sigma$ with $s$ → the t statistic

::::: columns
::: {.column width="60%"}
-   In practice, the population standard deviation $\sigma$ is rarely
    known.

-   Instead, we estimate population variability using the **sample
    standard deviation** $s$.

-   Substituting $s$ for $\sigma$ produces the [t statistic]{.keyword}.

$$
t = \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$

-   Because $s$ varies from sample to sample, this introduces **extra
    uncertainty** in the standardized mean.

-   As a result, the statistic follows a **t distribution** rather than
    the normal distribution.
:::

::: {.column width="40%"}
The t distribution

```{r}
#| label: fig-t-statistic-distribution
#| fig-cap: Student's t distribution for a small sample size, showing the distribution of the standardized mean when the population standard deviation is estimated from the sample.
#| fig-alt: Bell-shaped t distribution centered at zero, with the horizontal axis labeled t and the vertical axis labeled probability density.
p +
  geom_line(
    data = ~ filter(.x, distribution == "t"),
    linewidth = 1.5
  ) +
  labs(x = "t")
```
:::
:::::

## The [t distribution]{.keyword}

::::: columns
::: {.column width="60%"}
-   When the standard deviation is **estimated from the sample**, the
    standardized mean follows a **t distribution**.
-   The t distribution is similar to the normal distribution but has
    **heavier tails**.
-   Heavier tails reflect the additional uncertainty introduced when
    estimating $\sigma$ with $s$.
-   The shape of the t distribution depends on the **degrees of
    freedom**.
-   As sample size increases, the t distribution approaches the **normal
    distribution**.
:::

::: {.column width="40%"}
Comparing the Z and t distributions

```{r}
#| label: fig-z-vs-t-distribution
#| fig-cap: Comparison of the standard normal and Student's t distributions. The t distribution has heavier tails because estimating the population standard deviation adds uncertainty. Figure generated using R.
#| fig-alt: Line graph showing two bell-shaped curves centered at zero, one labeled Z and one labeled t. The t curve is slightly lower at the peak and higher in the tails than the Z curve.

lines <-
  tibble(
    distribution = "z",
    label = "Z",
    x = df |> 
      filter(distribution == "z") |> 
      slice(n()*.59) |> 
      pull(x),
    y = df |> 
      filter(distribution == "z") |> 
      slice(n()*.59) |> 
      pull(density),
    xend = x + .7,
    yend = y
  ) |> 
  add_row(
    distribution = "t",
    label = "t",
    x = df |> 
      filter(distribution == "t") |> 
      slice(n()*.82) |> 
      pull(x),
    y = df |> 
      filter(distribution == "t") |> 
      slice(n()*.82) |> 
      pull(density),
    xend = x + .4,
    yend = y + .05
  ) 

p +
  geom_line(linewidth = 1.5) +
  labs(x = "Z or t") +
  geom_segment(
    aes(x, y, xend = xend, yend = yend, color = NULL),
    data = lines
  ) +
  geom_label(
    aes(xend, yend, label = label, color = NULL),
    data = lines,
    linewidth = 0
  )
```
:::
:::::

## Degrees of Freedom

::::: columns
::: {.column width="50%"}
1.  The exact shape of the t distribution depends on the number of
    [degrees of freedom]{.keyword}.
2.  For a sample mean the degrees of freedom are $df = n - 1$.
3.  Larger degrees of freedom produce a distribution closer to normal.
:::

::: {.column width="50%"}
```{r}
#| label: fig-t-degrees-of-freedom
#| fig-cap: Student's t distributions with different degrees of freedom. As degrees of freedom increase, the t distribution approaches the normal distribution. Figure generated using R.
#| fig-alt: Line graph showing several bell-shaped curves centered at zero representing t distributions with different degrees of freedom, gradually becoming narrower and more similar to a normal distribution as degrees of freedom increase.

library(ggplot2)
library(dplyr)

# x values
x <- seq(-4, 4, length.out = 1000)

# degrees of freedom to illustrate
dfs <- c(1, 3, 10, 30)

# create data
df <- expand.grid(
  x = x,
  df = dfs
) |>
  mutate(
    density = dt(x, df),
    df = factor(df)
  )

# plot
ggplot(df, aes(x = x, y = density, color = df)) +
  geom_line(linewidth = 1.3) +
  scale_color_brewer(
    palette = "Dark2",
    name = "Degrees of freedom"
  ) +
  labs(
    title = "Effect of Degrees of Freedom on the t Distribution",
    x = "Standardized value (t)",
    y = "Probability density"
  ) +
  theme_classic(base_size = 18) +
  theme(
    legend.position = "right",
    axis.ticks = element_blank()
  )
```
:::
:::::

## What Are We Actually Estimating?

::::: columns
::: {.column width="50%"}
-   The quantity of interest is the population mean $\mu$.
-   The sample mean $\bar{x}$ provides an estimate of this parameter.
-   Because of sampling variability our estimate always contains
    uncertainty.
:::

::: {.column width="50%"}
```{r}
#| label: fig-what-are-we-estimating
#| fig-cap: A sample mean provides an estimate of the population mean. The confidence interval shows the uncertainty in that estimate.
#| fig-alt: Density curve representing a population distribution with a vertical line marking the population mean and a point with a horizontal confidence interval showing an estimate of the mean from a sample.

library(ggplot2)
library(dplyr)

set.seed(42)

# Simulated population
pop <- rnorm(100000, mean = 50, sd = 10)

# Sample
n <- 25
sample <- sample(pop, n)

xbar <- mean(sample)
s <- sd(sample)

# 95% CI
se <- s / sqrt(n)
t_crit <- qt(0.975, df = n - 1)

ci_low <- xbar - t_crit * se
ci_high <- xbar + t_crit * se

# Density data
df <- data.frame(x = pop)

ggplot(df, aes(x = x)) +
  geom_density(fill = "#56B4E9", alpha = 0.35, linewidth = 1) +
  geom_vline(xintercept = mean(pop), linewidth = 1.2) +
  annotate("point",
           x = xbar,
           y = 0,
           size = 4
  ) +
  
  annotate("segment",
           x = ci_low,
           xend = ci_high,
           y = 0,
           yend = 0,
           linewidth = 3
  ) +
  labs(
    x = "Value",
    y = "Density"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )
```
:::
:::::

## Confidence Intervals for the Mean

::::: columns
::: {.column width="50%"}
-   A [confidence interval]{.keyword} provides a range of plausible
    values for the population mean.
-   The interval accounts for uncertainty caused by sampling
    variability.
-   Is centered around the sample mean.
-   Extends outward by a margin of error determined by the t
    distribution.
-   Wider intervals indicate greater uncertainty in the estimate.
:::

::: {.column width="50%"}
```{r}
#| label: fig-confidence-interval-mean
#| fig-cap: A confidence interval for a population mean constructed from a sample. The interval is centered on the sample mean and extends outward by a margin of error determined by the t distribution.
#| fig-alt: Bell-shaped sampling distribution centered at a sample mean, with vertical lines marking the lower and upper bounds of a confidence interval and the sample mean labeled in the center.

library(ggplot2)
library(dplyr)

set.seed(123)

# Simulate a population
pop_mean <- 50
pop_sd <- 10

# Sample
n <- 25
sample <- rnorm(n, pop_mean, pop_sd)

xbar <- mean(sample)
s <- sd(sample)

# Standard error
se <- s / sqrt(n)

# t critical value
t_crit <- qt(0.975, df = n - 1)

# Confidence interval
ci_low <- xbar - t_crit * se
ci_high <- xbar + t_crit * se

# Sampling distribution curve (approximation)
x <- seq(xbar - 4*se, xbar + 4*se, length.out = 1000)
df <- tibble(
  x = x,
  density = dnorm(x, mean = xbar, sd = se)
)

ggplot(df, aes(x = x, y = density)) +
  geom_line(linewidth = 1.5) +
  geom_vline(xintercept = xbar, linewidth = 1.2) +
  geom_vline(xintercept = c(ci_low, ci_high), linetype = "dashed", linewidth = 1) +
  annotate(
    "segment",
    x = ci_low,
    xend = ci_high,
    y = 0,
    yend = 0,
    linewidth = 3
  ) +
  labs(
    title = "Confidence Interval for a Population Mean",
    x = "Possible values of the mean",
    y = "Probability density"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )
```
:::
:::::

## Calculating a Confidence Interval

::::: columns
::: {.column width="50%"}
-   Confidence intervals for means are calculated using the t
    distribution.
-   The interval combines the sample mean, the standard error, and a
    critical t value.
    -   The **critical value** comes from the *t distribution* and
        depends on the chosen **significance level** ( $\alpha$ ) and
        the **degrees of freedom** ( $df = n - 1$ ) for the sample.
-   The formula describes how far the interval extends from the sample
    mean.

$$
\bar{x} \pm t_{\alpha/2,df}\,SE
$$
:::

::: {.column width="50%"}
![Annotated diagram showing components of the confidence interval
formula. Source: Whitlock & Schluter
3e.](images/confidence-interval-formula.png){#fig-confidence-interval-formula
fig-alt="Equation for a confidence interval with labels identifying the sample mean, critical t value, and standard error."}
:::
:::::

::: notes
-   The 2SE rule of thumb from lecture 6 derives from this equation
    -   Uses a critical value of 2 (approximately 95% of sample means
        are within 2 SEs of the true mean).
:::

## Interpreting a Confidence Interval

::::: columns
::: {.column width="50%"}
-   A 95% confidence interval captures the true mean in 95% of repeated
    samples.
-   The interval represents uncertainty in estimating the population
    mean.
-   Narrow intervals indicate more precise estimates.
:::

::: {.column width="50%"}
```{r}
#| label: fig-many-confidence-intervals
#| fig-height: 7
#| fig-width: 8
#| fig-cap: "Repeated 95% confidence intervals for a population mean. Each horizontal line is a confidence interval from one sample; the vertical red line shows the true population mean. Most intervals include the true value, but some do not."

library(palmerpenguins)
library(dplyr)
library(ggplot2)

set.seed(123)

true_mean <- mean(penguins$body_mass_g, na.rm = TRUE)

n_intervals <- 20
sample_size <- 25

ci_data <- tibble(ci_id = 1:n_intervals) |>
  rowwise() |>
  mutate(
    sample = list(
      penguins |>
        filter(!is.na(body_mass_g)) |>
        slice_sample(n = sample_size)
    ),
    mean = mean(sample$body_mass_g),
    sd = sd(sample$body_mass_g),
    se = sd / sqrt(sample_size),
    lci = mean - 1.96 * se,
    uci = mean + 1.96 * se,
    contains_true = lci <= true_mean & uci >= true_mean
  ) |>
  ungroup()

ggplot(ci_data, aes(y = factor(ci_id))) +
  geom_segment(
    aes(x = lci, xend = uci, yend = factor(ci_id)),
    color = "#0072B2",
    linewidth = 1.2
  ) +
  geom_vline(
    xintercept = true_mean,
    color = "#D55E00",
    linewidth = 1.4
  ) +
  labs(
    title = "Repeated 95% CIs from random samples",
    subtitle = "Each interval based on a random sample of n = 25 penguins",
    caption = "Source: Palmer Penguins dataset (palmerpenguins package)",
    x = "Body mass (g)",
    y = "Confidence interval"
  ) +
  theme_minimal(base_size = 18)
```
:::
:::::

## Hypothesis Testing (Reminder)

::::: columns
::: {.column width="50%"}
-   Earlier lectures introduced the general framework of [hypothesis
    testing]{.keyword}.
-   We compare observed data with expectations under a null hypothesis.
-   If the result is unlikely under the null hypothesis we reject it.
:::

::: {.column width="50%"}
![Probability distribution of the number of right-handed toads expected
under the null hypothesis. The red tail regions represent outcomes at
least as extreme as the observed value, which together form the p-value.
Whitlock & Schluter
3e.](/lectures/slides/lecture-08-hypothesis-testing/images/toad-p-value.jpg){#fig-hypothesis-testing-rejection-regions
fig-alt="Histogram-like probability distribution for the number of right-handed toads, centered near the expected value of 9 under the null hypothesis. The observed value of 14 is marked on the right side, and red bars in both tails highlight outcomes as extreme or more extreme than the observation, representing the p-value."}
:::
:::::

## The [one-sample t test]{.keyword}

::::: columns
::: {.column width="50%"}
-   The one-sample t test evaluates whether a sample mean differs from a
    hypothesized mean $\mu_0$.
-   **Null hypothesis:** $H_0 : \mu = \mu_0$
-   **Alternative hypothesis:** $H_A : \mu \ne \mu_0$
-   The difference between $\bar{x}$ and $\mu_0$ is scaled by the
    standard error.
-   The resulting t statistic measures how extreme the observed
    difference is.

$$
t = \frac{\bar{x}-\mu_0}{SE}
$$
:::

::: {.column width="50%"}
![Student’s t distribution with 4 degrees of freedom showing the
rejection regions for a two-tailed test at α = 0.05. Each tail contains
2.5% of the probability, with critical values at 𝑡 = ±
2.78.](images/t-test-critical-values-two-tailed.jpg){#fig-t-test-critical-values-two-tailed
fig-alt="Bell-shaped t distribution with four degrees of freedom centered at zero. The horizontal axis ranges from −4 to 4 and the vertical axis shows probability density. The outer tails beyond −2.78 and 2.78 are shaded red, each labeled 2.5%, indicating the rejection regions for a two-tailed test at the 5% significance level."}
:::
:::::

## One-Sided and Two-Sided Tests

:::::: columns
::: {.column width="40%"}
-   A **two-sided test** evaluates whether the mean differs in either
    direction from $\mu_0$.
-   A **one-sided test** evaluates whether the mean is greater than or
    less than $\mu_0$.
-   Two-sided tests are typically used unless a directional hypothesis
    is justified.
:::

:::: {.column width="60%"}
::: {#fig-two-vs-one-tailed layout-ncol="2"}
![Two-tailed t test showing rejection regions in both tails (α = 0.05,
2.5% in each
tail).](images/t-test-critical-values-two-tailed.jpg){#fig-two-tailed
fig-align="center"}

![One-tailed t test showing the rejection region in a single tail (α =
0.05).](images/t-test-critical-values-one-tailed.jpg){#fig-one-tailed
fig-align="center"}

Comparison of rejection regions for two-tailed and one-tailed t tests.
In a two-tailed test, the significance level (α = 0.05) is split between
both tails of the distribution, producing two rejection regions of 2.5%
each. In a one-tailed test, the entire significance level is placed in a
single tail, producing a single rejection region of 5%.
:::
::::
::::::

## Assumptions of t Tests

::::: columns
::: {.column width="50%"}
-   Data should represent a random sample from the population.
-   The variable should be approximately normally distributed.
-   Moderate deviations from normality are often tolerated with larger
    samples.
:::

::: {.column width="50%"}
```{r}
#| label: fig-stream-chem-distributions
#| fig-cap: Distributions of stream chemistry variables from the [Luquillo LTER dataset](https://lter.github.io/lterdatasampler/reference/luq_streamchem.html). Some variables are approximately symmetric, while others are strongly skewed, illustrating the diversity of real-world data distributions.
#| fig-alt: Panel of small histograms with blue density curves showing the distributions of several stream chemistry variables. The shapes vary widely across panels, including roughly symmetric, right-skewed, and irregular distributions, illustrating that real-world variables often deviate from normality.

library(lterdatasampler)

luq_streamchem |> 
  select(-starts_with("sample"), -tdp) |> 
  tidyr::pivot_longer(!starts_with("sample")) |> 
  tidyr::drop_na() |> 
  ggplot(aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 9) +
  geom_density(color = "blue") +
  facet_wrap(~name, scales = "free") +
  labs(x = NULL, y = "Probability density") +
  theme_minimal(base_size = 18) +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
```
:::
:::::

## Sample Size and Interpretation

::::: columns
::: {.column width="50%"}
-   Larger samples reduce the standard error and narrow confidence
    intervals.
-   With large samples **even small differences** may become
    statistically significant.
-   Statistical significance does not always imply biological
    importance.
    -   Example: Average systolic blood pressure in a clinic sample (121
        mmHg) differs significantly from the recommended value of 120
        mmHg, but a 1 mmHg difference is unlikely to have meaningful
        consequences for patient health.
    -   This is why you always report effect size (means, differences in
        means), not just p-values
:::

::: {.column width="50%"}
```{r}
#| label: fig-ci-by-sample-size
#| fig-height: 7
#| fig-width: 14
#| fig-cap: "Repeated 95% confidence intervals for a population mean at two sample sizes (n = 10 and n = 25). Each horizontal line is one CI; the vertical red line is the true population mean. Larger samples produce narrower intervals. Data Source: Palmer Penguins."

library(tidyr)

set.seed(132)

true_mean <- mean(penguins$body_mass_g, na.rm = TRUE)

n_intervals <- 20
sample_sizes <- c(10, 25)

ci_data <- tidyr::expand_grid(
  n = sample_sizes,
  ci_id = 1:n_intervals
) |>
  rowwise() |>
  mutate(
    sample = list(
      penguins |>
        filter(!is.na(body_mass_g)) |>
        slice_sample(n = n)
    ),
    xbar = mean(sample$body_mass_g),
    s = sd(sample$body_mass_g),
    se = s / sqrt(n),
    lci = xbar - 1.96 * se,
    uci = xbar + 1.96 * se
  ) |>
  ungroup() |>
  mutate(
    n_facet = factor(paste0("n = ", n))
  )

ggplot(ci_data, aes(y = factor(ci_id))) +
  geom_segment(
    aes(x = lci, xend = uci, yend = factor(ci_id)),
    color = "#0072B2",
    linewidth = 1.2
  ) +
  geom_vline(
    xintercept = true_mean,
    color = "#D55E00",
    linewidth = 1.4
  ) +
  facet_wrap(~ n_facet, nrow = 1) +
  labs(
    x = "Body mass (g)",
    y = "CI number"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    strip.text = element_text(size = rel(1.5))
  )
```
:::
:::::

## Reporting Results

::::: columns
::: {.column width="50%"}
-   For a one-sample t-test:
    -   Report the sample mean, standard deviation, and confidence
        interval.
    -   Provide the t statistic, degrees of freedom, and p value.
-   For an estimated mean:
    -   Report the mean and confidence interval.
-   Results should be interpreted in the biological context of the
    study.
:::

::: {.column width="50%"}
Example:

> I tested whether the average systolic blood pressure in my sample
> differed from the clinical reference value of 120 mmHg and found a
> statistically significant result (one-sample t-test, t~39~=-2.45,
> two-sided p=0.019).
>
> The mean blood pressure was 117.2 mmHg, a difference of −2.8 mmHg from
> the reference value, with a 95% confidence interval of −5.1 to −0.5
> mmHg, indicating a small effect that is statistically detectable but
> unlikely to be clinically meaningful for individual patients.
:::
:::::
